---
title: "Seasonal Variations in Mortality Rates Among Shelter Residents in Toronto: Is Winter the Deadliest Season?"
author: 
  - Kevin Cai
thanks: "Code and data are available at: https://github.com/kevicai/toronto-sheltered-residents-deaths-analysis"
date: today
date-format: long
abstract: "This study explores Toronto winters' influence the morality rates among sheltered residents, utilizing data from 2007-2023. I explore on how cold weather impacts those in shelters by drawing upon previous research that highlights increased risks for homeless populations during colder months. I find that mortality rates in spring, summer, and fall are similar, while winter shows a slight increase compared to the other seasons. This finding suggests while homeless shelters do help to lower the risk of death in winter, further meaures should be taken to protect sheltered individuals against cold weather."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)

# read data
cleaned_data <- read_csv("../data/analysis_data/cleaned_data.csv")
summarized_data <- read_csv("../data/analysis_data/summarized_data.csv")
```

# Introduction

Homelessness is a persistent issue in Canada, with Toronto fostering to the largest population of homeless individuals in the country [@homelessStats]. On any given night, an estimated 10,000 people in Toronto are without a permanent home [@homelessStats], with many relying on Toronto shelters for safety and warmth. Despite the city of Toronto providing shelters for these homeless people, sheltered residents still face many health risks. 

Previous research has shown that homeless populations are at an increased risk of mortality by cold temperatures, particularly during the winter months [@winterAffets]. This paper aims to explore seasonal variations in mortality rates among shelter residents in Toronto between 2007 and 2024. By comparing mortality rates across winter, spring, summer, and fall, I seek to determine whether winter poses a disproportionately high risk for sheltered individuals. 

The remainder of this paper is structured as follows: @sec-data addresses the data used in the analysis, @sec-discussion disscusses my findings and limitations in the results.

# Data {#sec-data}

The data was extracted, cleaned, and analysed using the statistic programming software R [@citeR] and the R libaries `tidyverse` @citeTidyverse, `janitor` [@citeJanitor], `citeKnitr` [@citeKnitr]. 

## Overview of Dataset
To invesigate the affect of winter on sheltered residents in Toronto, the dataset "Deaths of Shelter Residents" [@datasetSource] collected by the City of Toronto’s Shelter and Support Services Division was used. This dataset has been systematically documenting the deaths of residents in its shelters since 2007. The number of deaths is recorded for each month, and each death is reported within 24 hours of identifying the deceased [@measurements]. These reportings apply to both current and recent shelter residents and no sensitive pravicy data that can identify the residents was recorded. The dataset measures deaths of male, female, and transgender/non-binary/two-Spirit sheltered residents. Similiar dataset are available by OpenDataToronto, including the number of deaths of people experiencing homelessness in general. However, the chosen dataset is the most relevant to the study of deaths of homeless people who have used Toronto shelters. 

I obtained the data using the OpenDataToronto R package [@citeOpenDataToronto] from OpenDataToronto. A sample of the cleaned dataset is shown in @tbl-cleaned. 

```{r, message = FALSE, echo = FALSE}
#| label: tbl-cleaned
#| tbl-cap: "Sample of Cleaned Deaths of Shelter Residents Data"
head(cleaned_data) |>
  kable(col.names = c("Year", "Season", "Deaths"))
```

The variables year and season denotes the year and season of death. The data originally contained the number of deaths for each month, but for the purpose of studying the affect of seasons the months are merged into seasons. Winter of each year includes December of its previous year, January and Feburary; spring includes March, April and May; Summer includes June, July, and August; Fall includes September, October, and November. The dataset does not contain the exact date of death, so the four seasons are divided three months each rather than the exact days for each season in Toronto. The original dataset includes months from January 2007 to August of 2024, but in the final dataset only months from March 2007 to Febuary 2024 are included to ensure that there are 17 occurrences of each season. Further, no invalid data was found in the original dataset and the cleaned dataset includes 68 rows. 

## Statistics of Dataset

```{r, message = FALSE, echo = FALSE}
#| label: tbl-summarized
#| tbl-cap: "Deaths of Shelter Residents Data"
head(summarized_data) |> 
  kable(col.names = c("Season", "Total Deaths", "Average Deaths"))
```

# Discussion {#sec-discussion}

# References


